---
title: "Astronaut database"
author: 
  - "Federico Di Franco"
  - "Henry Michel Iannella Jones"
  - "Francesco Pierri"
output: html_document
---

```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(tidyverse)
library(janitor)
library(knitr)
```

## 1. Data Loading

We begin by loading the data from the TidyTuesday repository. Using the direct URL ensures that the code is reproducible without relying on local files.

```{r load-data}
# Read in the data manually
url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-07-14/astronauts.csv"
astronauts_raw <- read_csv(url)

# Clean column names for easier typing and consistency
astronauts <- astronauts_raw %>% 
  clean_names()

# Check initial number of rows
nrow(astronauts)

# Display the first few rows to understand the data
head(astronauts)
```

## 2. Examining Data Structure
Next, we examine the dataset to understand its structure, dimensions, and the types of variables it contains. We use `str()` as recommended to avoid structure-related mistakes.

```{r check-structure}
# Check the structure and data types
str(astronauts)

# Show basic summary statistics
summary(astronauts)
```

Now, we will check for potential typos in categorical variables. This is important because typos can lead to incorrect groupings and analyses later on. We will look at the unique values in key categorical columns to identify any inconsistencies.

```{r check-categorical}
# Explore specific categorical columns for typos (e.g., 'pilot' vs 'Pilot')
# We sort alphabetically to make case-sensitivity issues stand out
typo_cols <- c("sex", "nationality", "military_civilian", "occupation")
for (col in typo_cols) {
  cat("\n--- Column:", col, "---\n")
  astronauts %>% 
    count(.data[[col]]) %>% 
    arrange(.data[[col]]) %>% 
    print(n = Inf) # Print all values to ensure no typo is missed
}
```

As we can see from the output above, there are some data entry inconsistencies. For example, in the `occupation` column we have both `"Pilot"` and `"pilot"`, or `"Space tourist"` and `"space tourist"`. We will address these case-sensitivity issues automatically in the cleaning phase.

## 3. Handling Missing Values and Inconsistencies
Before doing calculations, it is crucial to check for missing values and fix categorical 
inconsistencies. 

```{r check-missing}
# Count missing values per column
colSums(is.na(astronauts))

# Isolate rows that contain at least one missing value (NA) in any column
rows_with_na <- astronauts %>% 
  filter(if_any(everything(), is.na))
cat("\nTotal rows with at least one missing value:", nrow(rows_with_na), "\n")
```
We will inspect the rows with missing values to understand the nature of the missing data. This helps us decide how to handle them appropriately (e.g., whether to impute, drop, or categorize as "Unknown").

```{r inspect-missing}
# Display the first few of these rows to visually inspect the missing data
head(rows_with_na)
```

We will remove rows that are completely empty (if any). We make sure to check the number of rows before and after filtering to ensure we don't drop data unintentionally, even though we know there will be no NA numbers from the previous check, this makes our code robust for future additions as well.

```{r clean-data}
# Check rows before filtering
n_rows_before <- nrow(astronauts)
astronauts_clean <- astronauts %>% 
  # Remove rows where the key identifier 'number' is NA
  filter(!is.na(number)) 

# Check rows after filtering to ensure we didn't drop too much
n_rows_after <- nrow(astronauts_clean)
cat("Rows removed:", n_rows_before - n_rows_after, "\n")
```
Next, we will normalize all character columns to lowercase to automatically fix case typos (e.g., 'Pilot' -> 'pilot'). This is much safer and scalable than fixing typos manually one by one. We will also handle missing observations appropriately by replacing NA values in character columns with "Unknown". This way, we keep the numerical data of those astronauts intact while still acknowledging the missing information.
```{r fix-inconsistencies}
# Normalize all character columns to lowercase to automatically fix case typos (e.g., 'Pilot' -> 'pilot')
# This is much safer and scalable than fixing typos manually one by one.
astronauts_clean <- astronauts_clean %>%
  mutate(
    across(where(is.character), str_to_lower)
  )

# Handle missing observations appropriately
# Documenting our choice: We use "Unknown" for missing text data instead of dropping rows, 
# to keep numerical data of those astronauts intact.
astronauts_clean <- astronauts_clean %>%
  mutate(
    across(where(is.character), ~replace_na(.x, "Unknown"))
  )

# Verify that missing values have been handled
colSums(is.na(astronauts_clean))
```

In the end, we will verify that all typos have been fixed and that there are no unexpected missing values left in the cleaned dataset.

```{r verify-cleaning}
# Verify that all typos have been fixed by checking unique values again
for (col in typo_cols) {
  cat("\n--- Column:", col, "---\n")
  astronauts_clean %>%
    count(.data[[col]]) %>% 
    arrange(.data[[col]]) %>% 
    print(n = Inf) # Print all values to ensure no typo is missed
}
```

Now we can see that while the case-sensitivity issues have been resolved, there is still an inconsistency in the `occupation` column: some values contain the redundant prefix `"other (...)"` (e.g., `"other (journalist)"` and `"other (space tourist)"`). 

We want to extract just the specific role from inside the parentheses. We can do this cleanly using regular expressions with `str_replace()`:

```{r fix-occupation}
# Use regular expressions to extract the text inside 'other (...)'
astronauts_clean <- astronauts_clean %>%
  mutate(
    occupation = str_replace(occupation, "other \\((.*?)\\)", "\\1")
  )

# Verify the final cleaned occupation column
cat("\n--- Final 'occupation' column ---\n")
astronauts_clean %>%
  count(occupation) %>%
  arrange(occupation) %>%
  print(n = Inf)
```

## 4. Data Transformation
Finally, we transform the data types to make the dataset suitable for analysis, such as converting characters to factors where appropriate. This will help with any categorical analyses we want to perform later on.

```{r transform-data}
# Convert relevant character columns to factors
astronauts_clean <- astronauts_clean %>%
  mutate(
    sex = as.factor(sex),
    military_civilian = as.factor(military_civilian),
    nationality = as.factor(nationality),
    occupation = as.factor(occupation)
  )

# Verify the final structure
str(astronauts_clean)
```

The dataset `astronauts_clean` is now prepared and can be used for the next step: Exploratory Data Analysis.

# Exploratory Data Analysis

## 1. Description of Variables and Descriptive Statistics
The `astronauts_clean` dataset contains details about astronauts, their missions, and the time spent in space and on spacewalks (EVAs). We first calculate basic descriptive statistics for some key numerical variables.

```{r eda-summary}
# Summary of numerical features
astronauts_clean %>%
  summarise(
    n_obs = n(),
    mean_hours = mean(total_hrs_sum, na.rm = TRUE),
    median_hours = median(total_hrs_sum, na.rm = TRUE),
    sd_hours = sd(total_hrs_sum, na.rm = TRUE),
    max_hours = max(total_hrs_sum, na.rm = TRUE),
    mean_eva = mean(total_eva_hrs, na.rm = TRUE),
    max_eva = max(total_eva_hrs, na.rm = TRUE)
  )
```

```{r missing-check-eda}
# Missing values check specifically before the next transformations
astronauts_clean %>%
  summarise(across(everything(), ~ sum(is.na(.x)), .names = "na_{.col}"))
```

## 2. Creation of New Variables
As required, we will create at least 10 new columns (exactly 12) to enrich the dataset with meaningful calculated indicators, aggregated categories, and recoded variables. 

```{r new-variables}
astronauts_eda <- astronauts_clean %>%
  mutate(
    # 1. age_at_selection: age when selected
    age_at_selection = year_of_selection - year_of_birth,
    
    # 2. age_at_mission: age during the mission
    age_at_mission = year_of_mission - year_of_birth,
    
    # 3. decade_of_mission: grouping missions by decade
    decade_of_mission = as.factor(floor(year_of_mission / 10) * 10),
    
    # 4. career_duration: time between selection and mission
    career_duration = year_of_mission - year_of_selection,
    
    # 5. mission_duration_days: mission time in days
    mission_duration_days = hours_mission / 24,
    
    # 6. total_days_in_space: cumulative mission time in days
    total_days_in_space = total_hrs_sum / 24,
    
    # 7. is_us_or_ussr: superpower origin vs others 
    is_us_or_ussr = as.factor(if_else(str_detect(nationality, "u\\.s\\.|u\\.s\\.s\\.r"), "Superpower", "Other")),
    # Documenting our choice: We use str_detect() instead of exact matching (%in%) 
    # to correctly capture dual nationalities or mixed missions that contain a slash, 
    # such as "u.k./u.s." and "u.s.s.r/ukraine". This ensures no superpower is missed.
    
    # 8. has_eva: binary variable indicating if the mission had a spacewalk
    has_eva = as.factor(if_else(eva_hrs_mission > 0, "Yes", "No")),
    
    # 9. experience_level: distinguishing rookies from veterans
    experience_level = as.factor(case_when(
      mission_number == 1 ~ "Rookie", 
      TRUE ~ "Veteran"
    )),
    
    # 10. eva_percentage: percent of mission time spent on EVA (handling div by 0 just in case)
    eva_percentage = if_else(hours_mission > 0, (eva_hrs_mission / hours_mission) * 100, 0),
    
    # 11. mission_era: Historical era of spaceflight
    mission_era = factor(case_when(
      year_of_mission < 1975 ~ "Space Race",
      year_of_mission <= 2000 ~ "Shuttle & Mir",
      TRUE ~ "ISS Era"
    ), levels = c("Space Race", "Shuttle & Mir", "ISS Era")),
    
    # 12. eva_intensity_hrs_per_day: EVA hours normalized by mission duration
    eva_intensity_hrs_per_day = if_else(mission_duration_days > 0, eva_hrs_mission / mission_duration_days, 0)
  )

# Verification of the structure of the new dataset
glimpse(astronauts_eda)
```

## 3. Data Visualizations
Here are 12 meaningful charts to highlight the most important aspects of the data.

```{r custom-theme}
# --- Aesthetic Configuration ---
# Toggle this to 'TRUE' for the Dark Mode, or 'FALSE' for the Light Mode
use_dark_mode <- FALSE

if (use_dark_mode) {
  custom_theme <- theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "#1a1a2e", color = NA),
      panel.background = element_rect(fill = "#1a1a2e", color = NA),
      text = element_text(color = "white"),
      axis.text = element_text(color = "gray80"),
      plot.title = element_text(face = "bold", size = 16, color = "white"),
      plot.subtitle = element_text(face = "italic", color = "gray60", size = 12),
      panel.grid.major = element_line(color = "gray30", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      legend.background = element_rect(fill = "#1a1a2e", color = NA),
      legend.text = element_text(color = "white"),
      legend.title = element_text(color = "white", face = "bold"),
      legend.position = "top"
    )
  geom_color <- "#00b4d8"
} else {
  custom_theme <- theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      text = element_text(color = "black"),
      axis.text = element_text(color = "black"),
      plot.title = element_text(face = "bold", size = 16, color = "black"),
      plot.subtitle = element_text(face = "italic", color = "gray40", size = 12),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      legend.background = element_rect(fill = "white", color = NA),
      legend.text = element_text(color = "black"),
      legend.title = element_text(color = "black", face = "bold"),
      legend.position = "top"
    )
  geom_color <- "#03045e"
}

# Apply the theme globally so we don't have to repeat it in every chart
theme_set(custom_theme)
```

### Chart 1: Age Distribution During Missions
```{r chart-1-age}
# Why: To see the typical age of astronauts going to space.
# What: A histogram of age at mission.
# How it helps: Gives an idea of the physical requirements and career path timeline.
astronauts_eda %>%
  ggplot(aes(x = age_at_mission)) +
  geom_histogram(binwidth = 2, fill = geom_color, color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Age at Mission",
    subtitle = "Age of astronauts at the time of their launch",
    x = "Age",
    y = "Count"
  )
```

### Chart 2: Top 10 Nationalities
```{r chart-2-nationality}
# Why: To understand the geographic distribution of astronauts.
# What: A bar chart counting missions by the top 10 nationalities.
# How it helps: It highlights the historical dominance of specific countries in space exploration.
astronauts_eda %>%
  count(nationality, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(nationality, n), y = n, fill = nationality)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "viridis") +
  coord_flip() +
  labs(
    title = "Top 10 Nationalities by Number of Missions",
    subtitle = "Historically dominant space agencies clearly visible",
    x = "Nationality",
    y = "Number of Missions"
  )
```

### Chart 3: Mission Duration by Sex
```{r chart-3-sex-duration}
# Why: To check for gender disparities in mission assignments.
# What: A combination of a violin plot and jittered points for mission length.
# How it helps: Unlike a boxplot, this shows both the overall density shape and the exact concentration of individual missions (from short Shuttle flights to long ISS stays).
astronauts_eda %>%
  ggplot(aes(x = sex, y = mission_duration_days, fill = sex, color = sex)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    title = "Mission Duration (Days) by Sex",
    subtitle = "Showing both density distribution and individual data points",
    x = "Sex",
    y = "Duration (Days)"
  ) +
  theme(legend.position = "none")
```

### Chart 4: Total Space Hours over Time
```{r chart-4-trend-hours}
# Why: To see if missions are getting longer historically.
# What: A scatter plot of total hours vs mission year.
# How it helps: Highlights the evolution from short early flights to long-term ISS stays.
astronauts_eda %>%
  ggplot(aes(x = year_of_mission, y = total_hrs_sum, color = sex)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    title = "Total Space Hours vs. Year of Mission",
    subtitle = "Evolution from short flights to long-duration ISS stays",
    x = "Year of Mission",
    y = "Total Hours in Space",
    color = "Sex"
  )
```

### Chart 5: Missions per Year
```{r chart-5-missions-year}
# Why: To observe the pace of space exploration.
# What: A line plot showing the count of missions each year.
# How it helps: Shows peaks during the space race and shuttle eras.
astronauts_eda %>%
  count(year_of_mission) %>%
  ggplot(aes(x = year_of_mission, y = n)) +
  geom_line(color = geom_color, linewidth = 1.2, alpha = 0.8) +
  geom_point(color = geom_color, size = 2.5) +
  labs(
    title = "Number of Astronaut Missions per Year",
    subtitle = "Peaks indicate major programs like Apollo and Space Shuttle",
    x = "Year",
    y = "Count of Missions"
  )
```

### Chart 6: Experience Level of Astronauts
```{r chart-6-experience}
# Why: To check the ratio of rookies vs veterans sent to space.
# What: A bar chart of the experience level.
# How it helps: We can understand if space agencies rely more on fresh faces or experienced crews.
astronauts_eda %>%
  count(experience_level) %>%
  ggplot(aes(x = experience_level, y = n, fill = experience_level)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  labs(
    title = "Number of Missions by Astronaut Experience Level",
    subtitle = "Proportion of rookies vs returning veterans",
    x = "Experience Level",
    y = "Count"
  )
```

### Chart 7: Age at Selection: Military vs Civilian
```{r chart-7-selection-age}
# Why: To test hypotheses about career paths.
# What: A boxplot of age at selection grouped by military/civilian status.
# How it helps: Reveals whether military pilots are selected younger than civilians.
astronauts_eda %>%
  filter(!is.na(age_at_selection)) %>%
  ggplot(aes(x = military_civilian, y = age_at_selection, fill = military_civilian)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "cividis", begin = 0.4, end = 0.9) +
  labs(
    title = "Age at Selection: Military vs Civilian",
    subtitle = "Military personnel tend to be selected at slightly younger ages",
    x = "Background",
    y = "Age at Selection"
  )
```

### Chart 8: Career Duration vs Total Days in Space
```{r chart-8-career-duration}
# Why: To see if longer careers mean more time in space.
# What: Scatter plot comparing wait time (career duration) to total space time.
# How it helps: Identifies astronauts who waited a long time but flew short missions, vs highly utilized ones.
astronauts_eda %>%
  filter(career_duration >= 0) %>%
  ggplot(aes(x = career_duration, y = total_days_in_space)) +
  geom_point(alpha = 0.6, color = geom_color, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed") +
  labs(
    title = "Total Days in Space vs Career Duration before Mission",
    subtitle = "Checking if a longer wait leads to more time in space",
    x = "Years from Selection to Mission",
    y = "Total Days in Space"
  )
```

### Chart 9: Top Occupations
```{r chart-9-occupations}
# Why: To see what roles are most needed in space.
# What: Bar chart of the most frequent astronaut occupations.
# How it helps: Shows the shift from pilots to mission specialists and commanders.
astronauts_eda %>%
  count(occupation, sort = TRUE) %>%
  slice_head(n = 8) %>%
  ggplot(aes(x = reorder(occupation, n), y = n, fill = occupation)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "rocket") +
  coord_flip() +
  labs(
    title = "Top 8 Astronaut Occupations",
    subtitle = "Most frequent roles assigned during missions",
    x = "Occupation",
    y = "Count"
  )
```

### Chart 10: EVA Hours by Decade
```{r chart-10-eva-decade}
# Why: To examine the historical trend of spacewalking.
# What: Boxplot of EVA hours per mission, grouped by decade.
# How it helps: Illustrates how EVA (spacewalks) became more prominent in later decades (e.g., ISS construction).
astronauts_eda %>%
  ggplot(aes(x = decade_of_mission, y = eva_hrs_mission, fill = decade_of_mission)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Distribution of Mission EVA Hours by Decade",
    subtitle = "Historical increase in spacewalk duration and frequency",
    x = "Decade of Mission",
    y = "EVA Hours"
  )
```

### Chart 11: Astronaut Age by Space Era
```{r chart-11-age-by-era}
# Why: To see if modern astronauts are older on average than early space pioneers.
# What: A boxplot comparing age at mission across the three historical eras.
# How it helps: Shows the demographic shift from young test pilots to older scientists/specialists.
astronauts_eda %>%
  ggplot(aes(x = mission_era, y = age_at_mission, fill = mission_era)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "viridis") +
  labs(
    title = "Astronaut Age by Space Era",
    subtitle = "Modern ISS crews tend to be older than Space Race pioneers",
    x = "Space Era",
    y = "Age at Mission"
  )
```

### Chart 12: EVA Intensity Distribution
```{r chart-12-eva-intensity}
# Why: To understand the workload of spacewalks relative to mission length.
# What: A density plot of EVA hours per day of mission.
# How it helps: Identifies the typical concentration of extravehicular work during a flight.
astronauts_eda %>%
  filter(eva_intensity_hrs_per_day > 0) %>% # Focus only on missions with EVAs
  ggplot(aes(x = eva_intensity_hrs_per_day)) +
  geom_density(fill = geom_color, alpha = 0.7, color = "white") +
  labs(
    title = "Distribution of EVA Intensity",
    subtitle = "Average hours spent on spacewalks per day of mission",
    x = "EVA Hours per Day",
    y = "Density"
  )
```