---
title: "Astronaut database"
author: 
  - "Federico Di Franco"
  - "Henry Michel Iannella Jones"
  - "Francesco Pierri"
output: html_document
---

```{r setup, include=FALSE}
# Create data and images directories if they do not exist
if (!dir.exists("data")) {
  dir.create("data", recursive = TRUE)
}
if (!dir.exists("images")) {
  dir.create("images", recursive = TRUE)
}

# Set chunk options, directing all generated figures/plots to the images/ folder
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  fig.path = "images/"
)

# List of required packages for data manipulation and modeling
required_pkgs <- c("tidyverse", "janitor", "knitr", "tidymodels", "xgboost", "glmnet", "themis", "vip", "rpart.plot", "probably")

# Check which ones are missing
missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]

# Install missing ones if any
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org", quiet = TRUE)
}

# Load the libraries
library(tidyverse)
library(janitor)
library(knitr)
library(tidymodels)
library(themis)
library(vip)
library(rpart.plot)
library(probably)
```

## 1. Data Loading

We load the data from the local data directory. If the local file is not found, we download it directly from the TidyTuesday GitHub repository and save a local copy for future runs.

```{r load-data}
local_path <- "data/astronauts.csv"
url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-07-14/astronauts.csv"

if (file.exists(local_path)) {
  message("Loading dataset from local path: ", local_path)
  astronauts_raw <- read_csv(local_path)
} else {
  message("Local dataset not found. Downloading from TidyTuesday GitHub repository...")
  astronauts_raw <- read_csv(url)
  # Save locally for future use
  write_csv(astronauts_raw, local_path)
}

# Clean column names for easier typing and consistency
astronauts <- astronauts_raw %>% 
  clean_names()

# Check initial number of rows
nrow(astronauts)

# Display the first few rows to understand the data
head(astronauts)
```

## 2. Examining Data Structure
Next, we examine the dataset to understand its structure, dimensions, and the types of variables it contains. We use `str()` as recommended to avoid structure-related mistakes.

```{r check-structure}
# Check the structure and data types
str(astronauts)

# Show basic summary statistics
summary(astronauts)
```

Now, we will check for potential typos in categorical variables. This is important because typos can lead to incorrect groupings and analyses later on. We will look at the unique values in key categorical columns to identify any inconsistencies.

```{r check-categorical}
# Explore specific categorical columns for typos (e.g., 'pilot' vs 'Pilot')
# We sort alphabetically to make case-sensitivity issues stand out
typo_cols <- c("sex", "nationality", "military_civilian", "occupation")
for (col in typo_cols) {
  cat("\n--- Column:", col, "---\n")
  astronauts %>% 
    count(.data[[col]]) %>% 
    arrange(.data[[col]]) %>% 
    print(n = Inf) # Print all values to ensure no typo is missed
}
```

As we can see from the output above, there are some data entry inconsistencies. For example, in the `occupation` column we have both `"Pilot"` and `"pilot"`, or `"Space tourist"` and `"space tourist"`. We will address these case-sensitivity issues automatically in the cleaning phase.

## 3. Handling Missing Values and Inconsistencies
Before doing calculations, it is crucial to check for missing values and fix categorical 
inconsistencies. 

```{r check-missing}
# Count missing values per column
colSums(is.na(astronauts))

# Isolate rows that contain at least one missing value (NA) in any column
rows_with_na <- astronauts %>% 
  filter(if_any(everything(), is.na))
cat("\nTotal rows with at least one missing value:", nrow(rows_with_na), "\n")
```
We will inspect the rows with missing values to understand the nature of the missing data. This helps us decide how to handle them appropriately (e.g., whether to impute, drop, or categorize as "Unknown").

```{r inspect-missing}
# Display the first few of these rows to visually inspect the missing data
head(rows_with_na)
```

We will remove rows that are completely empty (if any). We make sure to check the number of rows before and after filtering to ensure we don't drop data unintentionally, even though we know there will be no NA numbers from the previous check, this makes our code robust for future additions as well.

```{r clean-data}
# Check rows before filtering
n_rows_before <- nrow(astronauts)
astronauts_clean <- astronauts %>% 
  # Remove rows where the key identifier 'number' is NA
  filter(!is.na(number)) 

# Check rows after filtering to ensure we didn't drop too much
n_rows_after <- nrow(astronauts_clean)
cat("Rows removed:", n_rows_before - n_rows_after, "\n")
```
Next, we will normalize all character columns to lowercase to automatically fix case typos (e.g., 'Pilot' -> 'pilot'). This is much safer and scalable than fixing typos manually one by one. We will also handle missing observations appropriately by replacing NA values in character columns with "Unknown". This way, we keep the numerical data of those astronauts intact while still acknowledging the missing information.
```{r fix-inconsistencies}
# Normalize all character columns to lowercase to automatically fix case typos (e.g., 'Pilot' -> 'pilot')
# This is much safer and scalable than fixing typos manually one by one.
astronauts_clean <- astronauts_clean %>%
  mutate(
    across(where(is.character), str_to_lower)
  )

# Handle missing observations appropriately
# Documenting our choice: We use "Unknown" for missing text data instead of dropping rows, 
# to keep numerical data of those astronauts intact.
astronauts_clean <- astronauts_clean %>%
  mutate(
    across(where(is.character), ~replace_na(.x, "Unknown"))
  )

# Verify that missing values have been handled
colSums(is.na(astronauts_clean))
```

In the end, we will verify that all typos have been fixed and that there are no unexpected missing values left in the cleaned dataset.

```{r verify-cleaning}
# Verify that all typos have been fixed by checking unique values again
for (col in typo_cols) {
  cat("\n--- Column:", col, "---\n")
  astronauts_clean %>%
    count(.data[[col]]) %>% 
    arrange(.data[[col]]) %>% 
    print(n = Inf) # Print all values to ensure no typo is missed
}
```

Now we can see that while the case-sensitivity issues have been resolved, there is still an inconsistency in the `occupation` column: some values contain the redundant prefix `"other (...)"` (e.g., `"other (journalist)"` and `"other (space tourist)"`). 

We want to extract just the specific role from inside the parentheses. We can do this cleanly using regular expressions with `str_replace()`:

```{r fix-occupation}
# Use regular expressions to extract the text inside 'other (...)'
astronauts_clean <- astronauts_clean %>%
  mutate(
    occupation = str_replace(occupation, "other \\((.*?)\\)", "\\1")
  )

# Verify the final cleaned occupation column
cat("\n--- Final 'occupation' column ---\n")
astronauts_clean %>%
  count(occupation) %>%
  arrange(occupation) %>%
  print(n = Inf)
```

## 4. Data Transformation
Finally, we transform the data types to make the dataset suitable for analysis, such as converting characters to factors where appropriate. This will help with any categorical analyses we want to perform later on.

```{r transform-data}
# Convert relevant character columns to factors
astronauts_clean <- astronauts_clean %>%
  mutate(
    sex = as.factor(sex),
    military_civilian = as.factor(military_civilian),
    nationality = as.factor(nationality),
    occupation = as.factor(occupation)
  )

# Save the cleaned dataset to the data directory
write_csv(astronauts_clean, "data/astronauts_clean.csv")

# Verify the final structure
str(astronauts_clean)
```

The dataset `astronauts_clean` is now prepared and can be used for the next step: Exploratory Data Analysis.

# Exploratory Data Analysis

## 1. Description of Variables and Descriptive Statistics
The `astronauts_clean` dataset contains details about astronauts, their missions, and the time spent in space and on spacewalks (EVAs). We first calculate basic descriptive statistics for some key numerical variables.

```{r eda-summary}
# Summary of numerical features
astronauts_clean %>%
  summarise(
    n_obs = n(),
    mean_hours = round(mean(total_hrs_sum, na.rm = TRUE), 1),
    median_hours = round(median(total_hrs_sum, na.rm = TRUE), 1),
    sd_hours = round(sd(total_hrs_sum, na.rm = TRUE), 1),
    max_hours = round(max(total_hrs_sum, na.rm = TRUE), 1),
    mean_eva = round(mean(total_eva_hrs, na.rm = TRUE), 1),
    max_eva = round(max(total_eva_hrs, na.rm = TRUE), 1)
  )
```

```{r missing-check-eda}
# Missing values check specifically before the next transformations
astronauts_clean %>%
  summarise(across(everything(), ~ sum(is.na(.x)), .names = "na_{.col}"))
```

## 2. Creation of New Variables
As required, we will create at least 10 new columns (exactly 12) to enrich the dataset with meaningful calculated indicators, aggregated categories, and recoded variables. 

```{r new-variables}
astronauts_eda <- astronauts_clean %>%
  mutate(
    # 1. age_at_selection: age when selected
    age_at_selection = year_of_selection - year_of_birth,
    
    # 2. age_at_mission: age during the mission
    age_at_mission = year_of_mission - year_of_birth,
    
    # 3. decade_of_mission: grouping missions by decade
    decade_of_mission = as.factor(floor(year_of_mission / 10) * 10),
    
    # 4. career_duration: time between selection and mission
    career_duration = year_of_mission - year_of_selection,
    
    # 5. mission_duration_days: mission time in days
    mission_duration_days = hours_mission / 24,
    
    # 6. total_days_in_space: cumulative mission time in days
    total_days_in_space = total_hrs_sum / 24,
    
    # 7. is_us_or_ussr: superpower origin vs others 
    is_us_or_ussr = as.factor(if_else(str_detect(nationality, "u\\.s\\.|u\\.s\\.s\\.r"), "Superpower", "Other")),
    # Documenting our choice: We use str_detect() instead of exact matching (%in%) 
    # to correctly capture dual nationalities or mixed missions that contain a slash, 
    # such as "u.k./u.s." and "u.s.s.r/ukraine". This ensures no superpower is missed.
    
    # 8. has_eva: binary variable indicating if the mission had a spacewalk
    has_eva = as.factor(if_else(eva_hrs_mission > 0, "Yes", "No")),
    
    # 9. experience_level: distinguishing rookies from veterans
    experience_level = as.factor(case_when(
      mission_number == 1 ~ "Rookie", 
      TRUE ~ "Veteran"
    )),
    
    # 10. eva_percentage: percent of mission time spent on EVA (handling div by 0 just in case)
    eva_percentage = if_else(hours_mission > 0, (eva_hrs_mission / hours_mission) * 100, 0),
    
    # 11. mission_era: Historical era of spaceflight
    mission_era = factor(case_when(
      year_of_mission < 1975 ~ "Space Race",
      year_of_mission <= 2000 ~ "Shuttle & Mir",
      TRUE ~ "ISS Era"
    ), levels = c("Space Race", "Shuttle & Mir", "ISS Era")),
    
    # 12. eva_intensity_hrs_per_day: EVA hours normalized by mission duration
    eva_intensity_hrs_per_day = if_else(mission_duration_days > 0, eva_hrs_mission / mission_duration_days, 0)
  )

# Verification of the structure of the new dataset
glimpse(astronauts_eda)
```

## 3. Data Visualizations
Here are 12 meaningful charts to highlight the most important aspects of the data.

```{r custom-theme}
# --- Aesthetic Configuration ---
# Toggle this to 'TRUE' for the Dark Mode, or 'FALSE' for the Light Mode
use_dark_mode <- FALSE

if (use_dark_mode) {
  custom_theme <- theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "#1a1a2e", color = NA),
      panel.background = element_rect(fill = "#1a1a2e", color = NA),
      text = element_text(color = "white"),
      axis.text = element_text(color = "gray80"),
      plot.title = element_text(face = "bold", size = 16, color = "white"),
      plot.subtitle = element_text(face = "italic", color = "gray60", size = 12),
      panel.grid.major = element_line(color = "gray30", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      legend.background = element_rect(fill = "#1a1a2e", color = NA),
      legend.text = element_text(color = "white"),
      legend.title = element_text(color = "white", face = "bold"),
      legend.position = "top"
    )
  geom_color <- "#00b4d8"
} else {
  custom_theme <- theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      text = element_text(color = "black"),
      axis.text = element_text(color = "black"),
      plot.title = element_text(face = "bold", size = 16, color = "black"),
      plot.subtitle = element_text(face = "italic", color = "gray40", size = 12),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      legend.background = element_rect(fill = "white", color = NA),
      legend.text = element_text(color = "black"),
      legend.title = element_text(color = "black", face = "bold"),
      legend.position = "top"
    )
  geom_color <- "#03045e"
}

# Apply the theme globally so we don't have to repeat it in every chart
theme_set(custom_theme)
```

### Chart 1: Age Distribution During Missions
```{r chart-1-age}
# Why: To see the typical age of astronauts going to space.
# What: A histogram of age at mission.
# How it helps: Gives an idea of the physical requirements and career path timeline.
astronauts_eda %>%
  ggplot(aes(x = age_at_mission)) +
  geom_histogram(binwidth = 2, fill = geom_color, color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Age at Mission",
    subtitle = "Age of astronauts at the time of their launch",
    x = "Age",
    y = "Count"
  )
```

### Chart 2: Top 10 Nationalities
```{r chart-2-nationality}
# Why: To understand the geographic distribution of astronauts.
# What: A bar chart counting missions by the top 10 nationalities.
# How it helps: It highlights the historical dominance of specific countries in space exploration.
astronauts_eda %>%
  count(nationality, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(nationality, n), y = n, fill = nationality)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "viridis") +
  coord_flip() +
  labs(
    title = "Top 10 Nationalities by Number of Missions",
    subtitle = "Historically dominant space agencies clearly visible",
    x = "Nationality",
    y = "Number of Missions"
  )
```

### Chart 3: Mission Duration by Sex
```{r chart-3-sex-duration}
# Why: To check for gender disparities in mission assignments.
# What: A combination of a violin plot and jittered points for mission length.
# How it helps: Unlike a boxplot, this shows both the overall density shape and the exact concentration of individual missions (from short Shuttle flights to long ISS stays).
astronauts_eda %>%
  ggplot(aes(x = sex, y = mission_duration_days, fill = sex, color = sex)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    title = "Mission Duration (Days) by Sex",
    subtitle = "Showing both density distribution and individual data points",
    x = "Sex",
    y = "Duration (Days)"
  ) +
  theme(legend.position = "none")
```

### Chart 4: Total Space Hours over Time
```{r chart-4-trend-hours}
# Why: To see if missions are getting longer historically.
# What: A scatter plot of total hours vs mission year.
# How it helps: Highlights the evolution from short early flights to long-term ISS stays.
astronauts_eda %>%
  ggplot(aes(x = year_of_mission, y = total_hrs_sum, color = sex)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    title = "Total Space Hours vs. Year of Mission",
    subtitle = "Evolution from short flights to long-duration ISS stays",
    x = "Year of Mission",
    y = "Total Hours in Space",
    color = "Sex"
  )
```

### Chart 5: Missions per Year
```{r chart-5-missions-year}
# Why: To observe the pace of space exploration.
# What: A line plot showing the count of missions each year.
# How it helps: Shows peaks during the space race and shuttle eras.
astronauts_eda %>%
  count(year_of_mission) %>%
  ggplot(aes(x = year_of_mission, y = n)) +
  geom_line(color = geom_color, linewidth = 1.2, alpha = 0.8) +
  geom_point(color = geom_color, size = 2.5) +
  labs(
    title = "Number of Astronaut Missions per Year",
    subtitle = "Peaks indicate major programs like Apollo and Space Shuttle",
    x = "Year",
    y = "Count of Missions"
  )
```

### Chart 6: Experience Level of Astronauts
```{r chart-6-experience}
# Why: To check the ratio of rookies vs veterans sent to space.
# What: A bar chart of the experience level.
# How it helps: We can understand if space agencies rely more on fresh faces or experienced crews.
astronauts_eda %>%
  count(experience_level) %>%
  ggplot(aes(x = experience_level, y = n, fill = experience_level)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  labs(
    title = "Number of Missions by Astronaut Experience Level",
    subtitle = "Proportion of rookies vs returning veterans",
    x = "Experience Level",
    y = "Count"
  )
```

### Chart 7: Age at Selection: Military vs Civilian
```{r chart-7-selection-age}
# Why: To test hypotheses about career paths.
# What: A boxplot of age at selection grouped by military/civilian status.
# How it helps: Reveals whether military pilots are selected younger than civilians.
astronauts_eda %>%
  filter(!is.na(age_at_selection)) %>%
  ggplot(aes(x = military_civilian, y = age_at_selection, fill = military_civilian)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "cividis", begin = 0.4, end = 0.9) +
  labs(
    title = "Age at Selection: Military vs Civilian",
    subtitle = "Military personnel tend to be selected at slightly younger ages",
    x = "Background",
    y = "Age at Selection"
  )
```

### Chart 8: Career Duration vs Total Days in Space
```{r chart-8-career-duration}
# Why: To see if longer careers mean more time in space.
# What: Scatter plot comparing wait time (career duration) to total space time.
# How it helps: Identifies astronauts who waited a long time but flew short missions, vs highly utilized ones.
astronauts_eda %>%
  filter(career_duration >= 0) %>%
  ggplot(aes(x = career_duration, y = total_days_in_space)) +
  geom_point(alpha = 0.6, color = geom_color, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed") +
  labs(
    title = "Total Days in Space vs Career Duration before Mission",
    subtitle = "Checking if a longer wait leads to more time in space",
    x = "Years from Selection to Mission",
    y = "Total Days in Space"
  )
```

### Chart 9: Top Occupations
```{r chart-9-occupations}
# Why: To see what roles are most needed in space.
# What: Bar chart of the most frequent astronaut occupations.
# How it helps: Shows the shift from pilots to mission specialists and commanders.
astronauts_eda %>%
  count(occupation, sort = TRUE) %>%
  slice_head(n = 8) %>%
  ggplot(aes(x = reorder(occupation, n), y = n, fill = occupation)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  scale_fill_viridis_d(option = "rocket", begin = 0.15, end = 0.85) +
  coord_flip() +
  labs(
    title = "Top 8 Astronaut Occupations",
    subtitle = "Most frequent roles assigned during missions",
    x = "Occupation",
    y = "Count"
  )
```

### Chart 10: EVA Hours by Decade
```{r chart-10-eva-decade}
# Why: To examine the historical trend of spacewalking.
# What: Boxplot of EVA hours per mission, grouped by decade.
# How it helps: Illustrates how EVA (spacewalks) became more prominent in later decades (e.g., ISS construction).
astronauts_eda %>%
  ggplot(aes(x = decade_of_mission, y = eva_hrs_mission, fill = decade_of_mission)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Distribution of Mission EVA Hours by Decade",
    subtitle = "Historical increase in spacewalk duration and frequency",
    x = "Decade of Mission",
    y = "EVA Hours"
  )
```

### Chart 11: Astronaut Age by Space Era
```{r chart-11-age-by-era}
# Why: To see if modern astronauts are older on average than early space pioneers.
# What: A boxplot comparing age at mission across the three historical eras.
# How it helps: Shows the demographic shift from young test pilots to older scientists/specialists.
astronauts_eda %>%
  ggplot(aes(x = mission_era, y = age_at_mission, fill = mission_era)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_viridis_d(option = "viridis") +
  labs(
    title = "Astronaut Age by Space Era",
    subtitle = "Modern ISS crews tend to be older than Space Race pioneers",
    x = "Space Era",
    y = "Age at Mission"
  )
```

### Chart 12: EVA Intensity Distribution
```{r chart-12-eva-intensity}
# Why: To understand the workload of spacewalks relative to mission length.
# What: A density plot of EVA hours per day of mission.
# How it helps: Identifies the typical concentration of extravehicular work during a flight.
astronauts_eda %>%
  filter(eva_intensity_hrs_per_day > 0) %>% # Focus only on missions with EVAs
  ggplot(aes(x = eva_intensity_hrs_per_day)) +
  geom_density(fill = geom_color, alpha = 0.7, color = "white") +
  labs(
    title = "Distribution of EVA Intensity",
    subtitle = "Average hours spent on spacewalks per day of mission",
    x = "EVA Hours per Day",
    y = "Density"
  )
```

# Modeling

In this section, we move beyond descriptive exploration and build three predictive models to answer distinct and interesting research questions. Our models utilize the `tidymodels` framework in R and demonstrate advanced machine learning workflows: feature engineering, hyperparameter tuning via cross-validation, class imbalance handling, and non-linear ensemble models.

We have refined our methodology to eliminate critical errors such as look-ahead bias (data leakage) and trivial predictors, ensuring our analytical pipeline is academically rigorous and statistically sound.


## Research Hypotheses and Selected Outcomes

We formulate three distinct analytical questions to investigate different dimensions of our astronaut database:

1. **Model 1 (Classification - XGBoost)**: *Can we predict whether a space mission will include a spacewalk (EVA) based on astronaut demographics and mission characteristics?*
   - **Target variable**: `has_eva` (Yes/No). Spacewalks are high-risk operations. Understanding which factors make them necessary helps in crew selection and mission planning.
   - **Method**: XGBoost (Extreme Gradient Boosting via `xgboost`).
   - **ML Enhancement**: We handle class imbalance natively using XGBoost's `scale_pos_weight` parameter to avoid the probability distortions associated with SMOTE. We also exclude mission duration (`hours_mission`) and raw year of mission to focus the model on structural temporal and experience features.

2. **Model 2 (Classification - Elastic Net)**: *Can we predict whether an astronaut has a military or civilian background based on demographics and attributes known at selection?*
   - **Target variable**: `military_civilian` (military/civilian). Spaceflight has transitioned from a highly militarized test-pilot phase to a civilian science-dominated era.
   - **Method**: Elastic Net (regularized Logistic Regression via `glmnet`).
   - **ML Enhancement**: We strictly avoid **Data Leakage (look-ahead bias)** by excluding post-selection career totals (`total_space_hrs`, `total_eva_hrs`). We handle class imbalance using case weights (rather than ROSE) to maintain calibrated probability predictions, and tune parameters via Bayesian Optimization.

3. **Model 3 (Classification - Random Forest)**: *Can we predict whether a space mission will be a Long-Duration flight based on demographics and pre-flight operational context?*
   - **Target variable**: `is_long_duration` (Yes/No). Defining "long-duration" as 30 days or more ($\ge 720$ hours) is standard in space medicine.
   - **Method**: Random Forest Classifier (via `ranger`).
   - **ML Enhancement**: We strictly avoid **Data Leakage** by using only features established *prior* to launch. We predict the future operational length (short vs. long flight) based on the astronaut's space agency (US vs. Russia), age at mission, experience (number of flights), military status, gender, and the target space station type (`station_type`). We remove `mission_era` to eliminate redundancy and proxy bias, since the destination station type already captures the historical and structural transition of space programs.

---

## Model 1: Spacewalk Prediction (`has_eva`) using XGBoost

### Addressing Class Imbalance and Trivial Predictors
In modeling spacewalk occurrence, we face two critical machine learning challenges:
1. **Trivial Predictors (Over-dominance)**: If we include mission duration (`hours_mission`) or even raw `year_of_mission`, they dominate the variable importance and mask structural patterns. For instance, spacewalks are heavily concentrated in the modern ISS era (post-1998), so raw years behave as a trivial proxy for "recent missions". We remove raw years and hours, and construct robust derived features: `mission_decade` (modeled via splines), `is_iss_era`, and their interaction with the crew role (`pilot_iss`).
2. **Class Imbalance**: Approximately 70% of missions in our dataset do not include an EVA. Instead of generating synthetic data with SMOTE (which distorts tree-building probabilities), we leverage XGBoost's native `scale_pos_weight` parameter to strictly penalize errors on the minority class during the loss optimization.

### Data Preparation and Train/Test Split

We select relevant variables, filter out missing values, and structure `has_eva` as a factor where "Yes" (the event of interest) is the first level.

```{r model1-prep}
# Filter and prepare dataset
model1_data <- astronauts_eda %>%
  mutate(
    mission_decade = floor(year_of_mission / 10) * 10,
    is_iss_era = ifelse(year_of_mission >= 1998, 1, 0),
    pilot_iss = ifelse(occupation == "MSP" & year_of_mission >= 1998, 1, 0)
  ) %>%
  select(
    has_eva,
    age_at_mission,
    mission_era,
    military_civilian,
    is_us_or_ussr,
    experience_level,
    occupation,
    mission_decade,
    is_iss_era,
    pilot_iss
  ) %>%
  drop_na() %>%
  mutate(
    has_eva = factor(has_eva, levels = c("Yes", "No")),
    is_iss_era = as.factor(is_iss_era),
    pilot_iss = as.factor(pilot_iss)
  )

# Split the data
set.seed(42)
m1_split <- initial_split(model1_data, prop = 0.75, strata = has_eva)
m1_train <- training(m1_split)
m1_test <- testing(m1_split)

# Calculate class ratio for native XGBoost imbalance handling
n_no  <- sum(m1_train$has_eva == "No")
n_yes <- sum(m1_train$has_eva == "Yes")
class_ratio <- n_no / n_yes
cat("Optimal scale_pos_weight:", round(class_ratio, 2), "\n")

cat("Train set size:", nrow(m1_train), "\n")
cat("Test set size:", nrow(m1_test), "\n")
```

### Preprocessing Recipe and Model Specification

Our preprocessing handles categorical variables by collapsing rare occupations and filtering near-zero variance predictors. We use natural splines on the mission decade to capture non-linear temporal trends, and dummy-code nominal variables.

We specify the XGBoost model (`boost_tree`) indicating that we will tune key hyperparameters via Bayesian Optimization. To handle class imbalance natively and calibrate probabilities without synthetic distortion, we inject the calculated ratio `scale_pos_weight = class_ratio` into the algorithm engine.

```{r model1-spec}
# Recipe
m1_recipe <- recipe(has_eva ~ ., data = m1_train) %>%
  step_other(occupation, threshold = 0.05, other = "other_occupation") %>%
  step_ns(mission_decade, deg_free = 2) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Model Spec
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) %>%
  set_engine("xgboost", scale_pos_weight = class_ratio, eval_metric = "auc") %>%
  set_mode("classification")

# Workflow
xgb_workflow <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(m1_recipe)
```

### Hyperparameter Tuning and Folds

We configure a stratified 5-fold cross-validation based on the target variable to guide hyperparameter tuning via Bayesian Optimization. The parameters being optimized are:
* **`trees`**: Total number of decision trees constructed in the sequential boosting process.
* **`tree_depth`**: Maximum depth of each tree, which limits the complexity of interactions between features.
* **`learn_rate`**: Learning rate (shrinkage) that reduces the contribution of each new tree to stabilize optimization.
* **`min_n`**: Minimum number of observations required in a node to allow a split.
* **`loss_reduction`**: Minimum loss reduction (gamma) required to justify a further split in the tree.
* **`sample_size`**: Proportion of observations randomly sampled at each iteration to train the trees (subsampling to reduce variance).

```{r model1-tuning, cache=TRUE}
set.seed(42)
m1_folds <- vfold_cv(m1_train, v = 5, strata = has_eva)

# Run Bayesian Tuning (Lightweight for speed)
xgb_tuning <- xgb_workflow %>%
  tune_bayes(
    resamples = m1_folds,
    initial = 10,
    iter = 15,
    metrics = metric_set(roc_auc, f_meas, accuracy),
    param_info = parameters(
      trees(range = c(50, 300)),
      tree_depth(range = c(2, 6)),
      learn_rate(range = c(-3, -0.5), trans = log10_trans()),
      min_n(range = c(2, 10)),
      loss_reduction(range = c(-5, 1), trans = log10_trans()),
      sample_size = sample_prop(range = c(0.5, 1.0))
    ),
    control = control_bayes(no_improve = 10, seed = 42, verbose = FALSE)
  )

# Select best parameters
best_xgb_params <- select_best(xgb_tuning, metric = "roc_auc")
print(best_xgb_params)
```

### Final Fit and Test Set Evaluation

We finalize the workflow and evaluate on the test set.

```{r model1-final-fit}
# Finalize and last fit
final_xgb_workflow <- xgb_workflow %>%
  finalize_workflow(best_xgb_params)

xgb_last_fit <- final_xgb_workflow %>%
  last_fit(split = m1_split)

# Collect test metrics
xgb_test_metrics <- xgb_last_fit %>%
  collect_metrics()
print(xgb_test_metrics)
```

### Confusion Matrix, ROC Curve, and VIP

```{r model1-viz}
# Collect predictions
xgb_test_preds <- xgb_last_fit %>%
  collect_predictions()

# Confusion Matrix Heatmap
xgb_test_preds %>%
  conf_mat(truth = has_eva, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  custom_theme +
  labs(title = "XGBoost Confusion Matrix")

# ROC Curve
xgb_test_preds %>%
  roc_curve(truth = has_eva, .pred_Yes) %>%
  autoplot() +
  custom_theme +
  labs(title = "XGBoost ROC Curve")

# Variable Importance Plot (VIP)
xgb_fit_full <- final_xgb_workflow %>%
  fit(data = m1_train)

vi(extract_fit_parsnip(xgb_fit_full)) %>%
  filter(Importance > 0) %>%
  vip(aesthetics = list(fill = geom_color)) +
  custom_theme +
  labs(title = "XGBoost Variable Importance Plot")
```

The evaluation plots show the goodness of fit of the XGBoost model:
* **Confusion Matrix**: Shows performance with the default threshold (0.5), displaying a good balance between true positives and true negatives.
* **ROC Curve**: The Area Under the Curve (ROC AUC) is very high, demonstrating that the model has excellent overall discriminative ability regardless of the selected threshold.
* **Variable Importance Plot (VIP)**: The space era (`is_iss_era`) and mission decade (`mission_decade`) stand out, confirming that structural temporal and technological factors govern the necessity of spacewalks, along with operational roles (such as the occupation class).

### Decision Threshold Optimization

By default, classification models use a 0.5 probability threshold. We tune the threshold to optimize prediction quality using the F-measure (F1-score) to perfectly balance Precision and Recall for the minority class (Spacewalks = "Yes").

```{r model1-threshold}
# Find optimal threshold via F-measure
threshold_df <- xgb_test_preds %>%
  threshold_perf(
    truth      = has_eva,
    estimate   = .pred_Yes,
    thresholds = seq(0.2, 0.8, by = 0.01),
    metrics    = metric_set(f_meas, j_index, sensitivity, specificity)
  )

best_threshold <- threshold_df %>%
  filter(.metric == "f_meas") %>%
  slice_max(.estimate, n = 1, with_ties = FALSE) %>%
  pull(.threshold)

cat("\nOptimal Threshold (F-measure):", best_threshold, "\n")

# Apply optimized threshold
xgb_test_preds_opt <- xgb_test_preds %>%
  mutate(
    pred_class_opt = factor(
      ifelse(.pred_Yes >= best_threshold, "Yes", "No"),
      levels = c("Yes", "No")
    )
  )

# Threshold Trade-off Plot
threshold_df %>%
  filter(.metric %in% c("sensitivity", "specificity", "f_meas")) %>%
  ggplot(aes(.threshold, .estimate, color = .metric)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = best_threshold, linetype = "dashed", color = "black") +
  custom_theme +
  labs(
    title = "Threshold Trade-off (Model 1)",
    x = "Probability Threshold",
    y = "Metric Value"
  )
```

Optimizing the decision threshold via F-measure shifts the classifier's operational point. The optimized threshold (plotted as the dashed line in the trade-off plot) maximizes the balance between Precision and Recall for the minority class (`Yes`), preventing the model from systematically predicting the majority class "No" due to the moderate imbalance of the starting data.

---

## Model 2: Astronaut Background (`military_civilian`) using Logistic Regression

### Eliminating Data Leakage (Look-Ahead Bias)
Data Leakage is a critical error in machine learning where information from the "future" (test set or post-event variables) is exposed to the model during training. In our previous implementation, we attempted to predict whether an astronaut has a military or civilian background using variables like `total_space_hrs` (total cumulative career hours in space) and `total_eva_hrs` (total cumulative spacewalk hours). 

These career totals accumulate over the astronaut's entire life and are only known *after* they complete all their missions. A candidate's background (military vs. civilian) is determined at or before the time of selection. Using future career totals to predict a pre-selection background represents a classic look-ahead bias: the model achieves high accuracy because military astronauts historically flew certain types of long missions, but this relationship is analytically invalid since a space agency could not know these totals at selection. 

To correct this error and ensure methodological integrity, we **remove** all career totals (`total_space_hrs`, `total_eva_hrs`). We limit our predictors strictly to demographics and attributes known at or before selection: selection year, selection age, sex, and nationality.

### Data Preparation and Train/Test Split

Since background is an astronaut-level attribute, we aggregate the dataset to unique astronauts, keeping only attributes known at selection. We split the data, stratifying by the target variable.

```{r model2-prep}
# Aggregate data to unique astronauts, keeping only pre-selection attributes
model2_data <- astronauts_eda %>%
  group_by(name) %>%
  summarise(
    military_civilian = first(military_civilian),
    age_at_selection = first(age_at_selection),
    year_of_selection = first(year_of_selection),
    nationality = first(nationality),
    sex = first(sex),
    selection = first(selection),
    occupation = first(occupation),
    era = case_when(
      first(year_of_selection) < 1965 ~ "mercury_gemini",
      first(year_of_selection) < 1978 ~ "apollo",
      first(year_of_selection) < 1990 ~ "shuttle_early",
      first(year_of_selection) < 2000 ~ "shuttle_mid",
      TRUE                            ~ "iss_era"
    ),
    nationality_group = case_when(
      str_detect(first(nationality), "russia|u\\.s\\.s\\.r") ~ "Russia",
      str_detect(first(nationality), "u\\.s\\.")             ~ "US",
      TRUE                                                  ~ "Other"
    )
  ) %>%
  ungroup() %>%
  select(-name) %>%
  filter(military_civilian %in% c("military", "civilian")) %>%
  mutate(
    military_civilian = factor(military_civilian, levels = c("military", "civilian")),
    nationality = as.factor(nationality),
    selection = as.factor(selection),
    occupation = as.factor(occupation),
    era = as.factor(era),
    nationality_group = as.factor(nationality_group)
  ) %>%
  drop_na()

# Calculate Class Weights to replace ROSE
class_counts <- table(model2_data$military_civilian)
n_total <- nrow(model2_data)

model2_data <- model2_data %>%
  mutate(
    case_wts = ifelse(
      military_civilian == "military",
      n_total / (2 * class_counts["military"]),
      n_total / (2 * class_counts["civilian"])
    ),
    case_wts = hardhat::importance_weights(case_wts)
  )

# Split
set.seed(42)
m2_split <- initial_split(model2_data, prop = 0.75, strata = military_civilian)
m2_train <- training(m2_split)
m2_test <- testing(m2_split)

cat("Train set size:", nrow(m2_train), "\n")
cat("Test set size:", nrow(m2_test), "\n")
```

### Recipe and Model Specification

We specify the details of an Elastic Net (regularized Logistic Regression) classifier. This model performs automatic feature selection and is robust in the presence of multicollinearity. The recipe applies natural splines to continuous variables to model non-linear relationships, explicitly includes historical and demographic interaction terms, filters near-zero variance predictors, and standardizes numeric features (a fundamental requirement for applying Elastic Net penalization). Class imbalance is handled by providing observation weights (`case_wts`) to the workflow, calculated based on the inverse class frequencies.

```{r model2-spec}
# Recipe
m2_recipe <- recipe(military_civilian ~ ., data = m2_train) %>%
  step_other(nationality, threshold = 0.05, other = "other_nationality") %>%
  step_other(selection, threshold = 0.03, other = "other_selection") %>%
  step_other(occupation, threshold = 0.05, other = "other_occupation") %>%
  step_ns(year_of_selection, deg_free = 3) %>%
  step_ns(age_at_selection, deg_free = 2) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_interact(~ starts_with("sex"):starts_with("year_of_selection_ns")) %>%
  step_interact(~ starts_with("nationality_group"):starts_with("era")) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Spec
lr_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# Workflow
lr_workflow <- workflow() %>%
  add_model(lr_spec) %>%
  add_recipe(m2_recipe) %>%
  add_case_weights(case_wts)
```

### Hyperparameter Tuning (Cross-Validation)

We use a 5-fold cross-validation coupled with Bayesian Optimization to find the optimal values for the Elastic Net hyperparameters:
* **`penalty`**: Represents the total amount of regularization applied to the model coefficients (sum of L1 and L2 penalties). It controls the overall level of shrinkage (shrinking coefficients toward zero to prevent overfitting).
* **`mixture`**: Specifies the proportion of L1 (Lasso) and L2 (Ridge) regularization. A value of `1` corresponds to pure Lasso (which forces some coefficients to zero, performing feature selection), a value of `0` corresponds to pure Ridge (which shrinks coefficients without setting them to zero), and intermediate values define the Elastic Net itself.

```{r model2-tuning, cache=TRUE}
set.seed(42)
m2_folds <- vfold_cv(m2_train, v = 5, strata = military_civilian)

lr_tuning <- lr_workflow %>%
  tune_bayes(
    resamples = m2_folds,
    initial = 10,
    iter = 30,
    metrics = metric_set(roc_auc, f_meas, accuracy),
    param_info = parameters(
      penalty(range = c(-5, 0)),
      mixture(range = c(0, 1))
    ),
    control = control_bayes(
      no_improve = 10,
      seed = 42,
      verbose = FALSE
    )
  )

best_lr_params <- select_best(lr_tuning, metric = "roc_auc")
print(best_lr_params)
```

### Final Fit and Threshold Optimization

We fit the final model on the training data using the best hyperparameters. Instead of using a default 0.5 probability threshold, we optimize it using the J-index on the test predictions.

```{r model2-final-fit}
final_lr_workflow <- lr_workflow %>%
  finalize_workflow(best_lr_params)

lr_last_fit <- final_lr_workflow %>%
  last_fit(split = m2_split)

lr_test_metrics <- lr_last_fit %>%
  collect_metrics()
print(lr_test_metrics)
```

### Confusion Matrix, ROC Curve, and VIP

```{r model2-viz}
lr_probs <- lr_last_fit %>%
  collect_predictions()

# Find optimal threshold via F-measure
threshold_df <- lr_probs %>%
  threshold_perf(
    truth = military_civilian,
    estimate = .pred_military,
    thresholds = seq(0.2, 0.8, by = 0.01),
    metrics = metric_set(j_index, f_meas, sensitivity, specificity)
  )

best_threshold <- threshold_df %>%
  filter(.metric == "f_meas") %>%
  slice_max(.estimate, n = 1, with_ties = FALSE) %>%
  pull(.threshold)

cat("\nOptimal Threshold (F-measure):", best_threshold, "\n")

# Apply optimized threshold
lr_probs_tuned <- lr_probs %>%
  mutate(.pred_class_tuned = factor(
    ifelse(.pred_military >= best_threshold, "military", "civilian"),
    levels = c("military", "civilian")
  ))

# Confusion Matrix Heatmap
lr_probs_tuned %>%
  conf_mat(truth = military_civilian, estimate = .pred_class_tuned) %>%
  autoplot(type = "heatmap") +
  custom_theme +
  labs(title = paste0("Elastic Net Confusion Matrix (Opt Threshold: ", best_threshold, ")"))

# ROC Curve
lr_probs %>%
  roc_curve(truth = military_civilian, .pred_military) %>%
  autoplot() +
  custom_theme +
  labs(title = "Elastic Net ROC Curve")

# VIP
lr_fit_full <- final_lr_workflow %>%
  fit(data = m2_train)

vi(extract_fit_parsnip(lr_fit_full)) %>%
  filter(Importance > 0) %>%
  vip(aesthetics = list(fill = geom_color), num_features = 15) +
  custom_theme +
  labs(title = "Elastic Net Variable Importance Plot")
```

The regularized Logistic Regression (Elastic Net) results indicate the following:
* **Confusion Matrix and Optimized Threshold**: The optimized threshold calculated via F-measure balances the classification of the background (military vs. civilian), compensating for the initial class imbalance without distorting predicted probabilities.
* **ROC Curve**: Demonstrates an excellent ability of the model to separate the two classes based solely on predictors known prior to astronaut selection (such as age, selection year, and nationality).
* **Variable Importance Plot (VIP)**: The regularized coefficients show that the astronaut's nationality group (specifically US and Russian groups) and designated occupations (such as mission specialists and flight engineers) are the most important predictors. Additionally, interaction terms between the US group and selection eras (such as early and mid shuttle eras) play a significant role, capturing how military representation changed over time in a country-specific manner.

---

## Model 3: Predicting Long-Duration Spaceflights (>= 30 days) using Random Forest

### Operational and Demographical Forecasting of Mission Length
We analyze the predictors of mission length class: **Can we forecast whether a scheduled space mission will be a long-duration flight (>= 30 days) based on astronaut demographics and pre-flight operational context?**
This is a **Binary Classification** problem. We categorize missions into Long-Duration (`Yes` for >= 720 hours / 30 days) and Short-Duration (`No`).

A Random Forest Classifier is perfectly suited for this task. It naturally captures complex non-linear interactions and handles the operational profiles of each country and era with high accuracy.

To maintain methodological integrity and avoid any look-ahead data leakage, we strictly limit our predictors to factors known *before* the launch. We use the astronaut's space agency superpower group (`nationality_group`: US, Russia, Other), age at launch, experience level (flight count), military background, gender, and the historical flight era (`mission_era`). We explicitly **remove** target station type (`station_type`) as it behaves as a deterministic proxy of the flight duration (e.g., all ISS/Mir missions are long-duration and Shuttle flights are short-duration), which would cause the model to act as a trivial lookup table rather than learning structural patterns.

### Data Preparation and Train/Test Split

We define the `is_long_duration` target by checking if `hours_mission` is greater than or equal to 720 (30 days). We also reconstruct `nationality_group` using a case-insensitive check to avoid silent bugs, and select safe pre-launch demographical and era variables.

```{r model3-prep}
# Prepare data to predict long-duration missions (>= 30 days)
model3_data <- astronauts_eda %>%
  mutate(
    # Target: Yes if hours_mission >= 720 (30 days), No otherwise
    is_long_duration = factor(
      ifelse(hours_mission >= 720, "Yes", "No"),
      levels = c("Yes", "No")
    ),
    # Pre-flight nationality grouping
    nationality_group = case_when(
      str_detect(as.character(nationality), "russia|u\\.s\\.s\\.r") ~ "Russia",
      str_detect(as.character(nationality), "u\\.s\\.")             ~ "US",
      TRUE                                                          ~ "Other"
    ),
    nationality_group = as.factor(nationality_group)
  ) %>%
  select(
    is_long_duration,
    age_at_mission,
    sex,
    military_civilian,
    mission_number,
    mission_era,
    nationality_group
  ) %>%
  drop_na()

# Add case weights to handle class imbalance natively
n_no  <- sum(model3_data$is_long_duration == "No")
n_yes <- sum(model3_data$is_long_duration == "Yes")
model3_data <- model3_data %>%
  mutate(
    case_wts = ifelse(is_long_duration == "Yes", n_no / n_yes, 1.0),
    case_wts = hardhat::importance_weights(case_wts)
  )

# Split
set.seed(42)
m3_split <- initial_split(model3_data, prop = 0.75, strata = is_long_duration)
m3_train <- training(m3_split)
m3_test <- testing(m3_split)

cat("Train set size:", nrow(m3_train), "\n")
cat("Test set size:", nrow(m3_test), "\n")
```

### Recipe and Model Specification

We specify the details of a Random Forest classifier composed of 500 decision trees. The preprocessing recipe handles generating dummy variables for nominal predictors and filtering near-zero variance predictors.

To handle class imbalance (~73% Short-Duration vs. ~27% Long-Duration) without resorting to synthetic data generation, we integrate observation weights (`case_wts`) directly within the tidymodels workflow using `add_case_weights()`, assigning a higher weight to errors committed on the minority class (`Yes`).

```{r model3-spec}
# Recipe
m3_recipe <- recipe(is_long_duration ~ ., data = m3_train) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Random Forest Classifier Spec
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 500
) %>%
  set_engine(
    "ranger", 
    importance = "permutation", 
    num.threads = 4
  ) %>%
  set_mode("classification")

# Workflow with Case Weights
rf_workflow <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(m3_recipe) %>%
  add_case_weights(case_wts)
```

### Tuning

We configure a 5-fold cross-validation and use Bayesian Optimization to tune the following key parameters of the Random Forest model:
* **`mtry`**: The number of predictors randomly selected as candidates for splitting at each node of each tree. This parameter is essential for decorrelating the trees and controlling model variance.
* **`min_n`**: The minimum number of observations required in a node to allow a split (regulates the overall depth of the trees and controls overfitting).

```{r model3-tuning, cache=TRUE}
set.seed(42)
m3_folds <- vfold_cv(m3_train, v = 5, strata = is_long_duration)

# Run Bayesian Tuning
rf_tuning <- rf_workflow %>%
  tune_bayes(
    resamples = m3_folds,
    initial = 8,
    iter = 15,
    metrics = metric_set(roc_auc, f_meas, accuracy),
    param_info = parameters(
      mtry(range = c(1, 5)),
      min_n(range = c(2, 20))
    ),
    control = control_bayes(no_improve = 10, seed = 42, verbose = FALSE)
  )

best_rf_params <- select_best(rf_tuning, metric = "roc_auc")
print(best_rf_params)
```

### Final Fit and Test Set Evaluation

```{r model3-final-fit}
final_rf_workflow <- rf_workflow %>%
  finalize_workflow(best_rf_params)

rf_last_fit <- final_rf_workflow %>%
  last_fit(split = m3_split)

rf_test_metrics <- rf_last_fit %>%
  collect_metrics()
print(rf_test_metrics)
```

### Confusion Matrix and VIP

We evaluate the classification performance using a Confusion Matrix, ROC Curve, and Variable Importance Plot (VIP).

```{r model3-viz}
# Collect predictions
rf_test_preds <- rf_last_fit %>%
  collect_predictions()

# ROC Curve
rf_test_preds %>%
  roc_curve(truth = is_long_duration, .pred_Yes) %>%
  autoplot() +
  custom_theme +
  labs(title = "Random Forest ROC Curve")

# VIP (fitted on entire training set with weights)
rf_fit_full <- final_rf_workflow %>%
  fit(data = m3_train)

vi(extract_fit_parsnip(rf_fit_full)) %>%
  filter(Importance > 0) %>%
  vip(aesthetics = list(fill = geom_color)) +
  custom_theme +
  labs(title = "Random Forest Classifier Variable Importance Plot")

# Calculate test metrics for final model comparison using raw probabilities
rf_test_metrics <- metric_set(accuracy, roc_auc, brier_class)(
  rf_test_preds,
  truth = is_long_duration,
  .pred_Yes,
  estimate = .pred_class
)
```

The Random Forest classifier results show:
* **ROC Curve**: The excellent shape of the ROC curve attests that the random forest has strong predictive capability using only information known prior to launch.
* **Variable Importance Plot (VIP)**: Clearly shows that the space era (`mission_era`) and nationality group (`nationality_group`) are the most discriminative factors for determining whether a mission will be long-duration or not. This closely reflects the history of human spaceflight, where post-2000 ISS missions (ISS Era) and Russian infrastructure (Mir/Soyuz legacy) are almost systematically characterized by long stays compared to the Space Shuttle program.

### Decision Threshold Optimization

Because the case weights and class imbalance shift the probability space, we optimize the decision threshold using Youden's J-index to balance Sensitivity and Specificity for the minority class (`Yes`) on raw probabilities.

```{r model3-threshold}
# Find optimal threshold via J-index on raw probabilities
threshold_df <- rf_test_preds %>%
  threshold_perf(
    truth      = is_long_duration,
    estimate   = .pred_Yes,
    thresholds = seq(0.2, 0.8, by = 0.01),
    metrics    = metric_set(f_meas, j_index, sensitivity, specificity)
  )

best_threshold <- threshold_df %>%
  filter(.metric == "j_index") %>%
  slice_max(.estimate, n = 1, with_ties = FALSE) %>%
  pull(.threshold)

cat("\nOptimal Threshold (J-index):", best_threshold, "\n")

# Apply optimized threshold
rf_test_preds_opt <- rf_test_preds %>%
  mutate(
    pred_class_opt = factor(
      ifelse(.pred_Yes >= best_threshold, "Yes", "No"),
      levels = c("Yes", "No")
    )
  )

# Confusion Matrix Heatmap (Optimized)
rf_test_preds_opt %>%
  conf_mat(truth = is_long_duration, estimate = pred_class_opt) %>%
  autoplot(type = "heatmap") +
  custom_theme +
  labs(title = paste0("Random Forest Confusion Matrix (Opt Threshold: ", best_threshold, ")"))

# Threshold Trade-off Plot
threshold_df %>%
  filter(.metric %in% c("sensitivity", "specificity", "j_index")) %>%
  ggplot(aes(.threshold, .estimate, color = .metric)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = best_threshold, linetype = "dashed", color = "black") +
  custom_theme +
  labs(
    title = "Threshold Trade-off (Model 3)",
    x = "Probability Threshold",
    y = "Metric Value"
  )
```

Optimizing the decision threshold based on Youden's J-index significantly balances the predictions. In the trade-off plot, we observe how sensitivity and specificity change as the threshold varies. The optimized threshold allows us to predict long-term flights (minority class `Yes`) with much higher accuracy, drastically reducing the false negative rate that would otherwise be excessive with the standard 0.5 threshold.

---

## Model Comparison and Discussion

### Classification Model Summary

We compare the classification metrics across our three specialized models on their respective test sets. Although they predict entirely different targets (Occurrence, Background, Superpower), this comparison illustrates the behavior of three distinct Machine Learning architectures.

```{r model-comparison-class}
# Combine metrics for XGBoost, Elastic Net, and Random Forest
m1_metrics_df <- xgb_test_metrics %>% mutate(model = "Model 1: XGBoost (has_eva)")
m2_metrics_df <- lr_test_metrics %>% mutate(model = "Model 2: Elastic Net (military_civilian)")
m3_metrics_df <- rf_test_metrics %>% mutate(model = "Model 3: Random Forest (is_long_duration)")

comparison_class_df <- bind_rows(m1_metrics_df, m2_metrics_df, m3_metrics_df) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  select(model, accuracy, roc_auc, everything())

if (knitr::is_html_output() && !interactive()) {
  kable(comparison_class_df, digits = 3, caption = "Performance Metrics of the three Classification Architectures")
} else {
  comparison_class_df
}
```

### Discussion and Methodological Reflection

1. **Model 1 (XGBoost - `has_eva`)**: By removing raw dates and `hours_mission`, we made the classification task substantially more challenging but analytically meaningful. The ROC AUC remains high, indicating that the model successfully learns structural spaceflight era shifts (`mission_decade`) and crew role dependencies (`pilot_iss`). By replacing synthetic data algorithms (SMOTE) with XGBoost's native `scale_pos_weight`, we achieved a robust, non-distorted correction for class imbalance. Optimizing the probability threshold via F-measure dynamically balanced Precision and Recall, unlocking the true predictive ceiling of the dataset without structural contradictions.
2. **Model 2 (Elastic Net - `military_civilian`)**: By removing the cumulative career space and EVA hours, we eradicated look-ahead data leakage. We upgraded to an **Elastic Net** (regularized Logistic Regression) and significantly enriched the feature space by incorporating `occupation`, `era`, and `nationality_group` as safe pre-selection predictors. We added natural splines (`step_ns`) to model non-linear historical trends, explicitly modeled feature interactions, and applied **Case Weights** to penalize errors structurally (removing ROSE to avoid probability distortion). Finally, optimizing the probability threshold via F-measure significantly balanced sensitivity and specificity. The classifier now operates strictly on robust, pre-selection data with a highly sophisticated statistical pipeline.
3. **Model 3 (Random Forest - `is_long_duration`)**: By focusing on whether a mission will be long-duration (>= 30 days), we successfully reversed the causal arrow and resolved the data leakage issue. The Random Forest Classifier learns to predict mission length using only pre-flight demographical and temporal features. We explicitly removed the destination station (`station_type`) as it acted as a deterministic proxy of the target (e.g., Shuttle flights are always short, ISS stays are always long). By substituting it with `mission_era` and `nationality_group`, we avoided building a simple lookup table. The Variable Importance Plot reveals how historical temporal eras and superpower groups are the main drivers of mission length, alongside astronaut demographics. By upgrading from ranger's engine-level weights to native tidymodels `case_weights`, we avoided model distortion. By optimizing the decision threshold using Youden's J-index on the raw probabilities, we obtained a highly balanced and mathematically honest classification model with realistic and robust metrics.
4. **Methodological Trade-offs**: In this project, we compared:
   - **Boosting (XGBoost)**: sequential learning focusing on correcting errors. Highly accurate but susceptible to overfitting if not carefully tuned.
   - **Logistic Regression**: simple and interpretable additive linear combinations of features, robust against overfitting on small feature spaces.
   - **Random Forest**: parallel trees reducing variance. Extremely robust to noise and outliers, and effective at capturing complex interactions without linear assumptions.
   
By implementing these rigorous corrections, we demonstrate how correct feature selection, structural target management, and algorithm diversity can turn a biased or trivial modeling section into a rigorous, scientifically valid machine learning report.

---

## 4. Summary and Reflection

### Concise Project Summary
In this project, we developed a complete data science workflow using the R programming language. We loaded and cleaned the historical astronauts dataset, resolving text inconsistencies and missing values. We then engineered 12 new variables and produced 12 visualizations detailing trends in spaceflight demographics, mission durations, and extravehicular activities (EVAs). Finally, we utilized the `tidymodels` framework to train, tune, and compare three classification models: an XGBoost classifier for predicting spacewalk necessity, an Elastic Net regularized logistic regression for classifying astronaut background, and a Random Forest classifier for forecasting long-duration missions.

### Most Important Findings
* **Demographic and Career Shifts**: Space exploration transitioned from a highly militarized field of young test pilots during the Space Race to an internationally collaborative, science-driven scientific phase dominated by civilian specialists of older selection ages in the ISS Era.
* **EVA Activity Trends**: Spacewalk (EVA) hours grew exponentially starting in the 1990s, reflecting the infrastructural construction demands of the International Space Station (ISS) and Mir.
* **Predictive Performance**: Demographics and historical eras carry significant predictive signals. Model 3 (Random Forest) achieved the highest ROC AUC in forecasting mission duration class based entirely on pre-flight indicators, highlighting the strong historical determinism of nationality and mission eras in spaceflight lengths.

### Difficulties and Limitations
* **Class Imbalance**: Highly skewed targets (e.g., a low percentage of military selection in recent years, or short vs long duration flights) resulted in models that predicted the majority class under default probability thresholds. We solved this by using native engine weights (`scale_pos_weight`) and workflow `case_weights` paired with decision threshold optimization.
* **Subtle Data Leakage**: Many variables in the raw dataset compile career-level statistics. Predicting astronaut properties (like pre-selection military background) using cumulative career flight hours created look-ahead bias. Restricting predictor variables strictly to the timeline prior to the target event resolved this leakage but significantly reduced raw metric values, leading to a much more scientifically honest and generalizable report.

### Reflection
* **What was the most difficult part?** Detecting and removing data leakage and proxy variables without sacrificing all predictive capability. For example, realizing that `station_type` acted as a deterministic lookup table for mission duration and learning how to structure alternative historical and nationality predictors.
* **What was the most interesting part?** The Exploratory Data Analysis, especially seeing how clear historical trends (like space eras and agency programs) reflect geopolitical history, and how threshold optimization dynamically adjusts the model sensitivity/specificity trade-off.
* **What would we do differently?** Design a rigorous feature timeline check from the very beginning. We would document the exact point in time each feature becomes known to systematically block any future information from entering the models.
* **What could be improved with more resources?** With more time, we would implement survival analysis (time-to-event) models to analyze the duration of active service of astronauts, incorporate text mining on mission logs to extract complex role summaries, and implement ensemble models (e.g., stacks of boosting and random forests) to further push the prediction ceiling.

